---
title: "Scutigera kinematics across developmental instars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scutigera kinematics across developmental instars}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)


devtools::load_all(".")
#library(scutigeraKinDev)

library(ggplot2)
library(ggrepel)
library(data.table)
library(trajr)

```

## Overview

This vignette demonstrates how to go from raw Tracker `.trk` files of
*Scutigera* locomotion to summary statistics and visualisations of
kinematic development across instars using the `scutigeraKinDev`
package. The pipeline mirrors an initial ad-hoc analysis script that
parsed XML, built trajectories with `{trajr}`, computed body and leg
kinematics, and generated diagnostic plots. 

We will:

1. Discover and parse `.trk` files.  
2. Build per-trial trajectories and body-level summaries.  
3. Visualise speed, body length, and kinetic energy across instars.  
4. TODO: Optionally, quantify leg protraction/retraction timing and “snow angel” plots in the body frame. 

Replace `path_to_trk_dir` with your actual directory when running this locally.

## 1. Discover and parse `.trk` files

```{r discover}

trk_dir  <- "~/Dropbox/Scutigera/data-2026-05-18/"

trk_files <- discover_trk_files(trk_dir)

length(trk_files)
head(trk_files)

```

`parse_trk_meta()` extracts global metadata (frame rate, scaling, units, etc.) if needed. [file:1]

```{r meta}

meta <- parse_trk_meta(trk_files)
head(meta)

```

Next, we turn each `.trk` file into a structured trial object with trajectories for all point masses.

```{r process-trials, message=FALSE}

trials <- lapply(trk_files, process_trk_file)

length(trials)

```

Each element of `trials` is a list with `meta` and `coords`, where `coords[["a"]]` and `coords[["p"]]` are the head and tail markers used for body-level metrics. 

## 2. Body-level kinematics across instars

We first compute body length and center-of-body speed summaries for each trial.

```{r body-stats}

stats <- summarise_body(trials)
head(stats)

```

Add a few derived quantities (relative speed, kinetic energy) matching the original script. [file:1]

```{r body-derived}

stats[, speed_relative_mean := speed_mean / body_length_mean]

stats[, mass_kg   := mass * 1e-6]       # mg → kg
stats[, speed_mps := speed_median * 1e-3]  # mm/s → m/s
stats[, Ekin_nJ   := mass_kg * speed_mps^2 / 2 * 1e9]  # J → nJ

stats2 <- stats[!is.na(body_length_mean) & !is.na(mass)]
head(stats2)

```

### Mean speed by instar

```{r plot-speed-mean}

ggplot(stats, aes(x = gsub("instar_", "", instar), colour = id)) +
  geom_pointrange(
    aes(
      y    = speed_mean,
      ymin = speed_mean - speed_sd,
      ymax = speed_mean + speed_sd
    ),
    position = position_dodge(width = 0.2)
  ) +
  ggrepel::geom_text_repel(aes(y = speed_mean, label = id)) +
  xlab("Instar") +
  ylab("Mean speed (mm/s)") +
  theme_bw()

  
```

### Median speed and kinetic energy across instars

```{r plot-speed-median}
  
ggplot(stats, aes(x = instar)) +
  geom_jitter(
    aes(
      y    = speed_median,
      size = mass
    ),
    colour = "black",
    width  = 0.05
  ) +
  ggrepel::geom_text_repel(aes(y = speed_median, label = id),
                           size = 3, colour = "gray50") +
  xlab("Instar") +
  ylab("Median speed (mm/s)") +
  labs(size = "Mass (mg)") +
  guides(colour = "none") +
  theme_bw() +
  theme(legend.position = "bottom")

```

```{r plot-ekin}
  
ggplot(stats, aes(x = instar)) +
  geom_jitter(
    aes(
      y    = Ekin_nJ,
      size = mass
    ),
    colour = "black",
    width  = 0.05
  ) +
  ggrepel::geom_text_repel(aes(y = Ekin_nJ, label = id),
                           size = 3, colour = "gray50") +
  xlab("Instar") +
  ylab("Kinetic energy (nJ)") +
  scale_y_log10() +
  labs(
    size  = "Mass (mg)",
    title = "Kinetic energy in the horizontal plane"
  ) +
  guides(colour = "none") +
  theme_bw() +
  theme(legend.position = "bottom")

```

### Body length and mass vs instar

```{r plot-body-length}

ggplot(stats, aes(x = instar, y = body_length_mean, colour = id)) +
  geom_jitter(width = 0.2) +
  xlab("Instar") +
  ylab("Head–tail length (mm)") +
  theme_bw()
  
```

```{r plot-mass}
  
ggplot(stats, aes(x = instar, y = mass, colour = id)) +
  geom_jitter(width = 0.2) +
  ggrepel::geom_text_repel(aes(label = id)) +
  xlab("Instar") +
  ylab("Mass (mg)") +
  theme_bw()

```

### Body length vs mass (allometric fit)

```{r plot-body-mass-fit}

fit <- nls(
  body_length_mean ~ b * (mass)^p,
  data  = stats2,
  start = list(b = 1, p = 0.5)
  )

  pred_df <- data.frame(
  mass = seq(min(stats2$mass), max(stats2$mass), length.out = 100)
  )
  pred_df$fit <- predict(fit, newdata = pred_df)

  cf <- coef(fit)
  fit_txt <- sprintf("Fitted: b = %.2f, p = %.2f", cf[["b"]], cf[["p"]])

  ggplot() +
  geom_point(
    data = stats2,
    mapping = aes(x = mass, y = body_length_mean)
  ) +
  ggrepel::geom_text_repel(
    data = stats2,
    mapping = aes(
      x = mass,
      y = body_length_mean,
      label = id,
      alpha = 0.5
    ),
    size = 3
  ) +
  geom_line(
    data = pred_df,
    mapping = aes(x = mass, y = fit)
  ) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title    = "Nonlinear fit: body length ~ b * mass^p",
    subtitle = fit_txt,
    x        = "Mass (mg)",
    y        = "Head–tail length (mm)"
  ) +
  guides(alpha = "none") +
  theme_minimal() +
  theme(legend.position = "bottom")
  
```

## 3. Leg kinematics and body-frame “snow angel” plots

